Patient-Specific Neural Field Modeling from Mirco-Electrode Data

### 2010-07-22 ltx/data_paper/DataPaper.tex
moved figures to end of paper, started to edit M-step and add E-step

\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/LFPs.eps}
\end{center}
\caption{{\bf Local field potentials from the micro-electrode array}. Data recorded from the first 8 channels of the array. The red dotted line indicates the seizure onset and the blue dashed line indicates the seizure end.}
\label{fig:TimeSeries}
\end{figure}

@@

\subsection{Estimating Support For Connectivity Kernel}
The support for the connectivity kernel can be inferred by studying the correlation between consecutive observations of the neural field. 
\begin{itemize}
	\item it is difficult to tell what correlations are from the cross-talk due the the sensor kernel
	\item it is difficult because of the nonlinear function
	\item 
\end{itemize}

\section{Estimation Algorithm}
\subsection{State-Space Model}
The state-space model is given by
\begin{align}
 \mathbf x_{t+1}&=q(\mathbf x_t)\boldsymbol\theta+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r) \\
 \mathbf y_t&=\mathbf C \mathbf x_t+\boldsymbol \epsilon_t
\end{align}
and the nonlinear function, $q(\cdot)$, is defined as
\begin{equation}\label{eq:QmatrixForSigmapoints}
	q(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}',
\end{equation}
where
\begin{equation}\label{eq:DefPsi}
	\boldsymbol{\Psi}(\mathbf{r}') \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\psi}^{\top} (\mathbf{r}'-\mathbf{r})d\mathbf{r}},
\end{equation}
$ \boldsymbol e_t(\mathbf r)\sim \mathcal N(\mathbf 0,\boldsymbol\Sigma_e(\mathbf r,\mathbf r'))$ is temporally white and spatially colored disturbance and  $\boldsymbol\epsilon_t\sim \mathcal N(\mathbf 0,\boldsymbol\Sigma_{\epsilon})$ is a white noise process.

\subsection{$\mathcal Q$-function}
The goal of the M-step is to maximize the likelihood function $p(\mathbf Y;\boldsymbol \theta,\xi)$ given the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ over the observed variables $\mathbf Y$ and hidden variables $\mathbf X$ that is governed by parameters $\boldsymbol \theta$ and $\xi$. Unfortunately, it is often difficult to maximize the likelihood $p(\mathbf Y;\boldsymbol \theta,\xi)$ directly. However, by using the relationship between $p(\mathbf Y;\boldsymbol \theta,\xi)$ and $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ it is possible to compute a sequence of parameter estimates that converges on the maximum likelihood parameter estimate.

@@

The $\mathcal Q$-function, or lower bound on the likelihood, can be written in terms of joint distribution components as
\begin{eqnarray}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=&\mathbf E_{\boldsymbol (\theta',\xi')}[\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)]\nonumber\\
&=&\mathbf E_{\boldsymbol (\theta',\xi')}[\sum_{t=1}^{T}\ln p(\mathbf y_t|\mathbf x_t)+\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta)+\sum_{t=1}^{T}\ln p(\mathbf x_0)]

@@

\mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\mathbf E_{(\theta',\xi')}[\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta ,\xi)]+\mathrm{const}

@@

the constant term can be ignored since it is not affecting the drivative with respect to the parameters. 

@@

where $\alpha$ is the normalising constant. Taking $\log$ we get

@@

&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t

@@

taking trace and using invariant cyclic permutations property of the trace this distribution can be written
as

@@

&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathrm{tr}\left\lbrace \mathbf x_t \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\right\rbrace 

@@

we first linearise $q(\mathbf x_t)$ by taking the first-order truncated  Taylor-series expansion around the mean of the state at each time instant for the sigmoidal activation function, i.e.
 f(\boldsymbol \phi^\top(\mathbf r')\mathbf x_t) \approx f(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)+  \boldsymbol \phi^\top(\mathbf r')(\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)
where we can approximate $q(\mathbf x_t)$ as
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q(\mathbf {\hat x}_t)+\int_\Omega \boldsymbol{\Psi}(\mathbf{r}')\boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'
By substituting \eqref{eq:qTaylor} in \eqref{eq:Qfunctionintrace}  and taking the expectation leads to the following  $\mathcal Q$-function
\begin{eqnarray}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=&\beta+2\Xi_0\boldsymbol\theta+

@@

&-&\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-2\xi \Xi_3 \boldsymbol\theta - \xi^2 tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace\nonumber
where $\beta= \sum_{t=1}^T\mathbf E_{(\theta',\xi')}[ \ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_w^{-1}\mathbf x_{t+1}]$ is constant with respect to $\boldsymbol \theta$ and will disapear by differentiation, and

@@

>>>>

@@

\subsection{Estimating Support For Connectivity Kernel}
The support for the connectivity kernel can be inferred by studying the cross-correlation between consecutive observations of the neural field. 
\begin{itemize}
	\item it is difficult to tell what correlations are from the cross-talk due the the sensor kernel
	\item it is difficult because of the nonlinear function
	\item 
\end{itemize}

\section{Estimation Algorithm}
The state-space model is given by
\begin{align}
 \mathbf x_{t+1}&=q(\mathbf x_t)\boldsymbol\theta+\xi \mathbf x_t+\boldsymbol e_t(\mathbf r) \\
 \mathbf y_t&=\mathbf C \mathbf x_t+\boldsymbol \epsilon_t,
\end{align}
where the nonlinear function, $q(\cdot)$, is defined as
\begin{equation}\label{eq:QmatrixForSigmapoints}
	q(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) d\mathbf{r}'
\end{equation}
and
\begin{equation}\label{eq:DefPsi}
	\boldsymbol{\Psi}(\mathbf{r}') \triangleq T_s\boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi}(\mathbf{r})\boldsymbol{\psi}^{\top} (\mathbf{r}'-\mathbf{r})d\mathbf{r}},
\end{equation}
$ \boldsymbol e_t(\mathbf r)\sim \mathcal N(\mathbf 0,\boldsymbol\Sigma_e(\mathbf r,\mathbf r'))$ is temporally white and spatially colored disturbance and  $\boldsymbol\epsilon_t\sim \mathcal N(\mathbf 0,\boldsymbol\Sigma_{\epsilon})$ is a white noise process.

\subsection{E-Step}
The unscented Rauch-Tung-Striebel smoother (URTSS)~\cite{Sarkka2010} was used for the state estimation. The URTSS incorporates an unscented Kalman filter (UKF)~\cite{Julier1997, Merwe2003} in a forward step to estimate posterior states, $\hat{\mathbf x}_t^{f}$, followed by a backward step to compute the smoothed state estimates, $\hat{\mathbf x}_t^{b}$. The first and second order moments of the predicted state are captured by propagating the so-called sigma points through the state equation. An additional element to the URTSS algorithm was created to approximate $q(\mathbf{x}_t)$, which is required for the M-step.

This is accomplished by taking the first-order truncated Taylor-series expansion of $q(\mathbf x_t)$ around the mean of the state at each time instant for the sigmoidal activation function, i.e.
\begin{equation}
 f(\boldsymbol \phi^\top(\mathbf r')\mathbf x_t) \approx f(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)+  \boldsymbol \phi^\top(\mathbf r')(\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t)
\end{equation}
where we can approximate $q(\mathbf x_t)$ as
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q(\mathbf {\hat x}_t)+\int_\Omega \boldsymbol{\Psi}(\mathbf{r}')\boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t)f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'
\end{equation}

The algorithm for the modified URTSS can be seen in Figure~\ref{tab:URTSSAlgorithm}.

\subsection{M-Step}

\subsection{$\mathcal Q$-function}
The goal of the M-step is to maximise the likelihood function $p(\mathbf Y;\boldsymbol \theta,\xi)$ given the joint distribution $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ over the observed variables $\mathbf Y$ and hidden variables $\mathbf X$ that is governed by parameters $\boldsymbol \theta$ and $\xi$. Unfortunately, it is often not possible to maximise the likelihood $p(\mathbf Y;\boldsymbol \theta,\xi)$ directly. However, by using the relationship between $p(\mathbf Y;\boldsymbol \theta,\xi)$ and $p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)$ it is possible to compute a sequence of parameter estimates that converges on the maximum likelihood parameter estimate.

@@

The $\mathcal Q$-function, or lower-bound on the likelihood, can be written in terms of joint distribution components as
\begin{align}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=\mathbf E_{\boldsymbol (\theta',\xi')}\left[\ln p(\mathbf X,\mathbf Y;\boldsymbol \theta,\xi)\right]\nonumber\\
&=\mathbf E_{\boldsymbol (\theta',\xi')}\left[\sum_{t=1}^{T}\ln p(\mathbf y_t|\mathbf x_t)+\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta)+\sum_{t=1}^{T}\ln p(\mathbf x_0)\right],
\end{align}

@@

\mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')=\mathbf E_{(\theta',\xi')}\left[\sum_{t=1}^{T}\ln p(\mathbf x_t|\mathbf x_{t-1};\boldsymbol \theta ,\xi)\right]+\mathrm{const}

@@

the constant term can be ignored since it is not affecting the derivative with respect to the parameters. 

@@

where $\alpha$ is the normalising constant. By expanding the exponent and taking $\log$ we get

@@

&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\mathbf x_t.

@@

Taking trace and rearranging, using the invariant cyclic permutations property of the trace, this distribution can be written as

@@

&&-\boldsymbol\theta^\top q(\mathbf x_t)^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-2\xi \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}q(\mathbf x_t)\boldsymbol\theta-\xi^2\mathrm{tr}\left\lbrace \mathbf x_t \mathbf x_t^\top\boldsymbol\Sigma_e^{-1}\right\rbrace. 

@@

Now taking the expectation of the log likelihood function gives the $\mathcal Q$-function, which is the lower-bound that needs to be maximised to gain optimal parameter estimates. The $\mathcal Q$-function is written as  
\begin{align}
 \mathcal Q(\boldsymbol \theta,\xi,\boldsymbol\theta',\xi')&=\beta+2\Xi_0\boldsymbol\theta+

@@

&-\boldsymbol\theta^\top \Xi_2\boldsymbol\theta-2\xi \Xi_3 \boldsymbol\theta - \xi^2 tr \left\lbrace\Xi_4\boldsymbol\Sigma_e^{-1} \right\rbrace \tag{Voldermont}
\end{align}
\begin{align}
	\beta &= \sum_{t=1}^T\mathbf E_{(\theta',\xi')}\left[ \ln \alpha-\mathbf x_{t+1}^\top\boldsymbol\Sigma_w^{-1}\mathbf x_{t+1}\right] \\
\Xi_0&=\sum_{t=1}^T\left[ \mathbf{\hat x}_{t+1}^\top\boldsymbol\Sigma_e^{-1}q(\mathbf{\hat x}_t)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_t \boldsymbol\Sigma_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t) d\mathbf{r}'\right] \\	
\Xi_1&=\mathbf E[\sum_{t=1}^T\mathbf x_t\mathbf x_{t+1}^\top] \\
\Xi_2&=\sum_{t=1}^T\Lambda_t(\mathbf r')\mathbf P_t \tilde{\Lambda}_t(\mathbf r')
\end{align}
is constant with respect to $\boldsymbol \theta$ and will disappear by differentiation, and

@@

% all figures here
% ~~~~~~~~~~~~~~~~
\begin{figure}[!ht]
\begin{center}
\includegraphics{./Figures/LFPs.eps}
\end{center}
\caption{{\bf Local field potentials from the micro-electrode array}. Data recorded from the first 8 channels of the array. The red dotted line indicates the seizure onset and the blue dashed line indicates the seizure end.}
\label{fig:TimeSeries}
\end{figure}

% tables here
% ~~~~~~~~~~~
\renewcommand{\arraystretch}{1.7}
\begin{table}[!ht]
\caption{
\bf{Algorithm for the Unscented RTS Smoother}}
\begin{tabular}{|c|}\hline
\multicolumn{1}{|p{16cm}|}{\textbf{1.} Forward initialisation} \\ 
$\hat{\mathbf x}_0, \mathbf P_0$ \\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{2.} Forward iteration: for $t \in \left\lbrace 0,\cdots, T\right\rbrace $, calculate the sigma points $\mathcal X_{i,t}^f$ using equations \ref{eq:sigmapoints1}-\ref{eq:sigmapoints3} and propagate through equation~\ref{eq:QmatrixForSigmapoints}} \\
$\mathcal X_{i,t+1}^{f-}=Q(\mathcal X_{i,t}^f) \quad i=0, \dots, 2L$\\
\multicolumn{1}{|p{16cm}|}{Calculate the predicted state and the predicted covariance matrix} \\
$\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{f-} $ \\
$\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e$ \\ 
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman filter update equations} \\
$\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\varepsilon})^{-1}$\\
$\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-}) $\\
$\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-}$\\ 
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{3.} Backward initialisation}\\
$\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T$ \\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{4.} Backward iteration: for $t \in \left\lbrace T-1, \cdots, 0 \right\rbrace $ calculate the sigma points $\mathcal X_{i,t}^b$ and propagate them through equation \ref{eq:QmatrixForSigmapoints}}\\
$\mathcal X_{i,t+1}^{b-}=Q(\mathcal X_{i,t}^b) \quad i=0, \dots, 2L$\\
\multicolumn{1}{|p{16cm}|}{Calculate the predicted state and the predicted covariance matrix}\\
$\hat{\mathbf x}_{t+1}^{b-}=\sum_{i=0}^{2L} W_i^{(m)}\mathcal X_{i,t+1}^{b-}$\\
$\mathbf P_{t +1}^{b-}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t +1}^{b-})^\top+\boldsymbol \Sigma_e $\\
$\mathbf M_{t +1}=\sum_{i=0}^{2L} W_i^{(c)}(\mathcal X_{i,t}^{b-}-\hat{\mathbf x}_{t}^{f})(\mathcal X_{i,t+1}^{b-}-\hat{\mathbf x}_{t+1}^{b-})^\top$\\
\multicolumn{1}{|p{16cm}|}{Compute the smoother gain, the smoothed state and the smoothed covariance matrix}\\
$\mathbf S_t=\mathbf M_{t +1}\left[ \mathbf P_{t +1}^{b-}\right] ^{-1} $\\
$\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathbf S_t\left[\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{b-}\right]$\\
$\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathbf S_t\left[\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{b-} \right]\mathbf S_t^\top $\\
\hline
\end{tabular}
\begin{flushleft}This table shows the steps in the unscented Rauch-Tung-Striebel smoother algorithm. The steps are iterated 11 times for our state estimation procedure. The least squares algorithm is run after each iteration to update the parameter estimates.
\end{flushleft}
\label{tab:UKFAlgorithm}
\end{table}
\renewcommand{\arraystretch}{1}
